\textbf{TITLE}

### 2022-04-01 submission/manuscript.Rmd
manscript draft and figs

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
library(knitr)
opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
	print(x)

	if(is.list(x)){
		x <- unlist(x)
	}

	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
		}
	} else {
    	paste(x)      
	}
}
knitr::knit_hooks$set(inline=inline_hook)
```

\vspace{20mm}

@@

\vspace{20mm}

@@

\vspace{20mm}

@@

## Introduction (~250 words)  
- Gut microbiome community composition has proven useful as a resource for machine learning prediction of various diseases {examples}. 
- Amplicon sequencing of the 16S rRNA gene is a reliable tool for assessing the taxonomic composition of microbial communities. 
- Analysis of 16S rRNA sequence data generally relies on clustering of sequences based on similarity into operational taxonomic units (OTUs).
- However OTU clustering depends on the data in the dataset and the addition of new data may change the overall OTU clusters. 
- The unstable nature of OTU clustering complicates machine learning. When building a classification model, changes to the OTU clusters means you have to re-create the model. This can change the underlying model and be time consuming and resource intensive. 
- The ability to integrate new data into an existing model could allow for deployment of a single model that new data can be continually added to and predicted on.
- Recently Sovacool *et al* described a new method for fitting new data into existing OTU clusters {Kelly optifit 2022}. 
- While OptiFit works well to fit new sequence data and provide high quality OTU clusters, it is unknown if the use of OptiFit will have an impact on machine learning predictions.
- Here, we use OptiFit with a 16S rRNA sequence dataset consisting of normal and SRN samples to test how well new data integrated with OptiFit performs for prediction of SRN.

## Results (~700 words)
- Utilize public dataset with normal and SRN samples to compare prediction between OptiFit and OptiClust
- randomly split data into 80% training 20% test sets 100 times
- Processed the data with both algorithms - Figure 1
  - 1. Used traditional OptiClust method to cluster all data, then split into training and test set
  - 2. Used OptiClust on the 80% training set, then used OptiFit to fit the remaining 20%
- OptiFit produces similar quality OTU clusters based on MCC (supplement?)
- Used mikropml to train a model on the 80% training set for each data split, then predicted dx on the 20% test set
- Training performance almost identical (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693) - Figure 2
- Performance on the test set comparable (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693) - Figure 2
- Maybe expand on where optifit did better/worse?

## Discussion/Conclusions (~250 words)
- OptiFit works!
- future questions:
  - how much reference do you need?
  - does it work well for other situations?
    - what if we used a new dataset for test set instead of a subset of the full dataset?
    - other diseases?

## Materials and Methods
- 16S rRNA amplicon sequence data from 490 subjects {baxter}
  - 261 controls
  - 229 SRN
- created 100 random splits of the data (80% training, 20% test)
- preprocessed data with mothur v1.45
- two pathways  
  -opticlust:  
    - clustered all data together
    - split the shared file based on the random splits

  -optifit:  
    - split the data
    - opticlust on the 80% training set
    - optifit to fit the remaining 20% to the training set OTUs
- ML with mikropml package (version)
  - preprocessed training set and applied preprocessing to test set
    - correlated collapsed, removed nzv

## Acknowledgements
- funding

@@

@@

>>>>

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE, include=FALSE}
library(tidyverse)
library(knitr)
opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)

```

**Title Options:**  
Applying OptiFit: An Improved Method for Fitting Amplicon Sequences to Existing OTUs to Predict Colorectal Cancer  
Predicting colorectal cancer with an improved method for fitting amplicon sequences to existing OTUs  
Efficient microbiome-based machine learning prediction of colorectal cancer using OptiFit  
OptiFit streamlines microbiome-based machine learning prediction of colorectal cancer  
OptiFit enables integration of new amplicon sequence data to existing OTUs for prediction of colorectal cancer

\vspace{10mm}

@@

\vspace{10mm}

@@

@@

Gut microbiome community composition is useful as a resource for machine learning prediction of various diseases {}. Amplicon sequencing of the 16S rRNA gene is a reliable tool for assessing the taxonomic composition of microbial communities. Analysis of 16S rRNA sequence data generally relies on clustering of sequences based on similarity into operational taxonomic units (OTUs). However, OTU clustering depends on the data in the dataset and the addition of new data could change the overall OTU clusters. The unstable nature of OTU clustering complicates deployment of machine learning models since integration of additional data requires re-clustering all the data and re-training of the model. The ability to integrate new data into an existing model without re-clustering and re-training could allow for deployment of a single model that new data can be continually added to and predicted on. Recently Sovacool *et al* described a new method for fitting new data into existing OTU clusters {Kelly optifit 2022}. While OptiFit works well to fit new sequence data and provide high quality OTU clusters, it is unknown if the use of OptiFit will have an impact on machine learning predictions. Here, we use OptiFit with a 16S rRNA sequence dataset consisting of normal and SRN samples to test how well new data integrated with OptiFit performs for prediction of SRN.

To assess machine learning prediction performance using OptiFit, we downloaded a publicly available dataset of 16S rRNA amplicon sequences from stool samples. The data set includes samples from healthy subjects as well as subjects with screen-relevant neoplasia (SRN) consisting of advanced adenoma and carcinoma. To simulate a scenario where we build a model with a set of data and then add additional data, we randomly split the dataset where 80% of the data was clustered into OTUs and the remaining 20% was used integrated into the existing OTUs with the OptiFit algorithm (Figure 1A). To account for variation depending on the split of the data, the data was randomly split 100 times and the process repeated for each of the 100 data splits. For comparison, the standard process using OptiClust was conducted where all of the data was clustered into OTUs using the OptiClust algorithm and the resulting OTU abundance table was split into the testing and training set (Figure 1B).

```{r mcc,include=F}
merged_mcc <- read_csv("../data/learning/summary/merged_MCC.csv")

opticlust_mcc <- merged_mcc %>% 
  filter(algorithm == "opticlust") %>% 
  pull(mcc)

optifit_avg_mcc <- merged_mcc %>% 
  filter(algorithm == "optifit" & state=="combo") %>% 
  summarise(mean_mcc = mean(mcc))
```
One way to examine the quality of the OTU clusters is with the Matthews correlation coefficient (MCC) which looks at the similarity between all pairs of sequences and assess whether they are appropriately in the same OTU or not {Westcott 2017 mSphere OptiClust}. An MCC score of zero indicates low quality OTU clusters while a score of 1 indicates high quality OTU clusters. Since the data is only clustered once in the OptiClust pathway there is only one MCC score while the OptiFit method produces an MCC score for the OTU clusters from each data split. Overall the MCC scores were similar between OptiClust (MCC = `r round(opticlust_mcc,digits=3)`) and OptiFit (average MCC = `r round(optifit_avg_mcc,digits=3)`) indicating that OptiFit does a good job of appropriately integrating new sequences into the existing OTUs.

A potential problem to using OptiFit for machine learning prediction is that any sequences in the new data that do not map to the existing OTU clusters will be discarded resulting in a possible loss of information. To quantify whether this impacts model performance we used the taxonomic abundances of the training data from the OptiClust and OptiFit pathways to train a model to predict SRNs based on OTU abundances. This model was then deployed to predict the diagnosis classification of the held out test data. To compare model performance we calculated the area under the receiver operating characteristic curve (AUROC) for each data split on both the training data and the testing data. The model performance during training was equivalent between the two algorithms (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693, Figure 2A). Additionally, the performance on the test data was equivalent between the two algorithms (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693, Figure 2A,B) indicating that new data can be integrated into existing OTU clusters without impacting model performance.

Overall this analysis demonstrated that OptiFit can be used to fit new sequence data into existing OTU clusters and perform equally in predicting SRNs compared to clustering all of the sequence data together. The ability to integrate new data into existing OTUs enables the deployment of a single machine learning model based on microbiome composition that new data can be predicted on. These results are based on a single dataset and disease. Further analysis is needed to determine if the process works well for data collected or processed with different methodology or prediction of other conditions.

## Materials and Methods
***Data Set.*** Raw 16S rRNA amplicon sequence data isolated from human stool samples was downloaded from NCBI Sequence Read Archive (accession no. SRP062005) {baxter}. This data set contains stool samples from a total of N subjects, however after preprocessing to screen for sequence quality and subsample to 10,000 reads per sample, 490 samples remained. For this analysis, samples from subjects identified in the metadata as normal, high risk normal, or adenoma were categorized as "normal" while samples from subjects identified as advanced adenoma or carcinoma were categorized as "screen relevant neoplasia" (SRN). The resulting data set consisted of 261 normal samples and 229 SRN samples.

***Data Processing.*** The full dataset was pre-processed with mothur (v1.45) using the SILVA reference database (v132) {silva} to join forward and reverse reads, merge duplicate reads, align to the reference, pre-cluster, remove chimeras, assign taxonomy, and remove non-bacterial reads following the Schloss Lab MiSeq standard operating procedure described on the mothur website (https://mothur.org/wiki/miseq_sop/). 100 splits of the 490 samples were generated where 80% of the samples (392 samples) were randomly assigned to the training set and the remaining 20% (98 samples) were assigned to the test set. Using 100 splits of the data accounts for the variation that may be observed depending on the samples that are in the training or test sets. Each samples was in the training set an average of N times and the test set and average of N times.

The data was processed through two pathways. First, the standard pathway using the OptiClust algorithm. In this pathway, all of the data was clustered together to generate OTUs and the resulting abundance tables were split into the training and testing sets. In the second pathway, the pre-processed data was split into the training and testing sets. The training set was clustered into OTUs, then the test set was fit to the OTUs of the training set using the OptiFit algorithm. The OptiFit algorithm was run with method open so that any sequences that didn't map to the existing OTU clusters would form new OTUs. Any OTUs that were not in the training set were removed prior to machine learning. For both pathways all data was sub-sampled to 10,000 reads per sample.

***Machine Learning.*** Machine learning using Random Forest was conducted with the R package mikrompl (v XXXX) {} to predict the diagnosis (SRN or normal) for the samples in the test set for each data split. The training set was preprocessed to normalize values (scale/center), collapse colinear features, and remove features with zero-variance. The preprocessing from the training set was then applied to the test set. P values comparing model performance were calculated as previously described {}. The averaged ROC curves were plotted by taking the average and standard deviation of the sensitivity at each specificity value.

***Code Availability.*** all code used in this analysis is available at https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2021

## Acknowledgements
(funding)

@@

![](../exploratory/figures/figure1B.png)
![](../exploratory/figures/figure1Apng.png)

@@

![](../exploratory/figures/Figure2.png)

### 2022-04-01 submission/manuscript.tex
manscript draft and figs

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\title{\textbf{TITLE}}
\author{}
\date{\vspace{-2.5em}}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={TITLE},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

@@

@@

\vspace{20mm}

@@

\hypertarget{introduction-250-words}{%
\subsection{Introduction (\textasciitilde250
words)}\label{introduction-250-words}}

\begin{itemize}
\tightlist
\item
  Gut microbiome community composition has proven useful as a resource
  for machine learning prediction of various diseases \{examples\}.
\item
  Amplicon sequencing of the 16S rRNA gene is a reliable tool for
  assessing the taxonomic composition of microbial communities.
\item
  Analysis of 16S rRNA sequence data generally relies on clustering of
  sequences based on similarity into operational taxonomic units (OTUs).
\item
  However OTU clustering depends on the data in the dataset and the
  addition of new data may change the overall OTU clusters.
\item
  The unstable nature of OTU clustering complicates machine learning.
  When building a classification model, changes to the OTU clusters
  means you have to re-create the model. This can change the underlying
  model and be time consuming and resource intensive.
\item
  The ability to integrate new data into an existing model could allow
  for deployment of a single model that new data can be continually
  added to and predicted on.
\item
  Recently Sovacool \emph{et al} described a new method for fitting new
  data into existing OTU clusters \{Kelly optifit 2022\}.
\item
  While OptiFit works well to fit new sequence data and provide high
  quality OTU clusters, it is unknown if the use of OptiFit will have an
  impact on machine learning predictions.
\item
  Here, we use OptiFit with a 16S rRNA sequence dataset consisting of
  normal and SRN samples to test how well new data integrated with
  OptiFit performs for prediction of SRN.
\end{itemize}

\hypertarget{results-700-words}{%
\subsection{Results (\textasciitilde700
words)}\label{results-700-words}}

\begin{itemize}
\tightlist
\item
  Utilize public dataset with normal and SRN samples to compare
  prediction between OptiFit and OptiClust
\item
  randomly split data into 80\% training 20\% test sets 100 times
\item
  Processed the data with both algorithms - Figure 1

  \begin{itemize}
  \item
    \begin{enumerate}
    \def\labelenumi{\arabic{enumi}.}
    \tightlist
    \item
      Used traditional OptiClust method to cluster all data, then split
      into training and test set
    \end{enumerate}
  \item
    \begin{enumerate}
    \def\labelenumi{\arabic{enumi}.}
    \setcounter{enumi}{1}
    \tightlist
    \item
      Used OptiClust on the 80\% training set, then used OptiFit to fit
      the remaining 20\%
    \end{enumerate}
  \end{itemize}
\item
  OptiFit produces similar quality OTU clusters based on MCC
  (supplement?)
\item
  Used mikropml to train a model on the 80\% training set for each data
  split, then predicted dx on the 20\% test set
\item
  Training performance almost identical (OptiClust median CV AUC 0.694,
  OptiFit median CV AUC 0.693) - Figure 2
\item
  Performance on the test set comparable (OptiClust median CV AUC 0.694,
  OptiFit median CV AUC 0.693) - Figure 2
\item
  Maybe expand on where optifit did better/worse?
\end{itemize}

\hypertarget{discussionconclusions-250-words}{%
\subsection{Discussion/Conclusions (\textasciitilde250
words)}\label{discussionconclusions-250-words}}

\begin{itemize}
\tightlist
\item
  OptiFit works!
\item
  future questions:

  \begin{itemize}
  \tightlist
  \item
    how much reference do you need?
  \item
    does it work well for other situations?

    \begin{itemize}
    \tightlist
    \item
      what if we used a new dataset for test set instead of a subset of
      the full dataset?
    \item
      other diseases?
    \end{itemize}
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  16S rRNA amplicon sequence data from 490 subjects \{baxter\}

  \begin{itemize}
  \tightlist
  \item
    261 controls
  \item
    229 SRN
  \end{itemize}
\item
  created 100 random splits of the data (80\% training, 20\% test)
\item
  preprocessed data with mothur v1.45
\item
  two pathways\\
  -opticlust:

  \begin{itemize}
  \tightlist
  \item
    clustered all data together
  \item
    split the shared file based on the random splits
  \end{itemize}

  -optifit:

  \begin{itemize}
  \tightlist
  \item
    split the data
  \item
    opticlust on the 80\% training set
  \item
    optifit to fit the remaining 20\% to the training set OTUs
  \end{itemize}
\item
  ML with mikropml package (version)

  \begin{itemize}
  \tightlist
  \item
    preprocessed training set and applied preprocessing to test set

    \begin{itemize}
    \tightlist
    \item
      correlated collapsed, removed nzv
    \end{itemize}
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  funding
\end{itemize}

@@

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={TITLE},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

@@

\title{\textbf{TITLE}}
\author{}
\date{\vspace{-2.5em}}

@@

\textbf{Title Options:}\\
Applying OptiFit: An Improved Method for Fitting Amplicon Sequences to
Existing OTUs to Predict Colorectal Cancer\\
Predicting colorectal cancer with an improved method for fitting
amplicon sequences to existing OTUs\\
Efficient microbiome-based machine learning prediction of colorectal
cancer using OptiFit\\
OptiFit streamlines microbiome-based machine learning prediction of
colorectal cancer\\
OptiFit enables integration of new amplicon sequence data to existing
OTUs for prediction of colorectal cancer

\vspace{10mm}

@@

Gut microbiome community composition is useful as a resource for machine
learning prediction of various diseases \{\}. Amplicon sequencing of the
16S rRNA gene is a reliable tool for assessing the taxonomic composition
of microbial communities. Analysis of 16S rRNA sequence data generally
relies on clustering of sequences based on similarity into operational
taxonomic units (OTUs). However, OTU clustering depends on the data in
the dataset and the addition of new data could change the overall OTU
clusters. The unstable nature of OTU clustering complicates deployment
of machine learning models since integration of additional data requires
re-clustering all the data and re-training of the model. The ability to
integrate new data into an existing model without re-clustering and
re-training could allow for deployment of a single model that new data
can be continually added to and predicted on. Recently Sovacool \emph{et
al} described a new method for fitting new data into existing OTU
clusters \{Kelly optifit 2022\}. While OptiFit works well to fit new
sequence data and provide high quality OTU clusters, it is unknown if
the use of OptiFit will have an impact on machine learning predictions.
Here, we use OptiFit with a 16S rRNA sequence dataset consisting of
normal and SRN samples to test how well new data integrated with OptiFit
performs for prediction of SRN.

To assess machine learning prediction performance using OptiFit, we
downloaded a publicly available dataset of 16S rRNA amplicon sequences
from stool samples. The data set includes samples from healthy subjects
as well as subjects with screen-relevant neoplasia (SRN) consisting of
advanced adenoma and carcinoma. To simulate a scenario where we build a
model with a set of data and then add additional data, we randomly split
the dataset where 80\% of the data was clustered into OTUs and the
remaining 20\% was used integrated into the existing OTUs with the
OptiFit algorithm (Figure 1A). To account for variation depending on the
split of the data, the data was randomly split 100 times and the process
repeated for each of the 100 data splits. For comparison, the standard
process using OptiClust was conducted where all of the data was
clustered into OTUs using the OptiClust algorithm and the resulting OTU
abundance table was split into the testing and training set (Figure 1B).

One way to examine the quality of the OTU clusters is with the Matthews
correlation coefficient (MCC) which looks at the similarity between all
pairs of sequences and assess whether they are appropriately in the same
OTU or not \{Westcott 2017 mSphere OptiClust\}. An MCC score of zero
indicates low quality OTU clusters while a score of 1 indicates high
quality OTU clusters. Since the data is only clustered once in the
OptiClust pathway there is only one MCC score while the OptiFit method
produces an MCC score for the OTU clusters from each data split. Overall
the MCC scores were similar between OptiClust (MCC = 0.895) and OptiFit
(average MCC = 0.892) indicating that OptiFit does a good job of
appropriately integrating new sequences into the existing OTUs.

A potential problem to using OptiFit for machine learning prediction is
that any sequences in the new data that do not map to the existing OTU
clusters will be discarded resulting in a possible loss of information.
To quantify whether this impacts model performance we used the taxonomic
abundances of the training data from the OptiClust and OptiFit pathways
to train a model to predict SRNs based on OTU abundances. This model was
then deployed to predict the diagnosis classification of the held out
test data. To compare model performance we calculated the area under the
receiver operating characteristic curve (AUROC) for each data split on
both the training data and the testing data. The model performance
during training was equivalent between the two algorithms (OptiClust
median CV AUC 0.694, OptiFit median CV AUC 0.693, Figure 2A).
Additionally, the performance on the test data was equivalent between
the two algorithms (OptiClust median CV AUC 0.694, OptiFit median CV AUC
0.693, Figure 2A,B) indicating that new data can be integrated into
existing OTU clusters without impacting model performance.

Overall this analysis demonstrated that OptiFit can be used to fit new
sequence data into existing OTU clusters and perform equally in
predicting SRNs compared to clustering all of the sequence data
together. The ability to integrate new data into existing OTUs enables
the deployment of a single machine learning model based on microbiome
composition that new data can be predicted on. These results are based
on a single dataset and disease. Further analysis is needed to determine
if the process works well for data collected or processed with different
methodology or prediction of other conditions.

@@

\textbf{\emph{Data Set.}} Raw 16S rRNA amplicon sequence data isolated
from human stool samples was downloaded from NCBI Sequence Read Archive
(accession no. SRP062005) \{baxter\}. This data set contains stool
samples from a total of N subjects, however after preprocessing to
screen for sequence quality and subsample to 10,000 reads per sample,
490 samples remained. For this analysis, samples from subjects
identified in the metadata as normal, high risk normal, or adenoma were
categorized as ``normal'' while samples from subjects identified as
advanced adenoma or carcinoma were categorized as ``screen relevant
neoplasia'' (SRN). The resulting data set consisted of 261 normal
samples and 229 SRN samples.

\textbf{\emph{Data Processing.}} The full dataset was pre-processed with
mothur (v1.45) using the SILVA reference database (v132) \{silva\} to
join forward and reverse reads, merge duplicate reads, align to the
reference, pre-cluster, remove chimeras, assign taxonomy, and remove
non-bacterial reads following the Schloss Lab MiSeq standard operating
procedure described on the mothur website
(\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
samples were generated where 80\% of the samples (392 samples) were
randomly assigned to the training set and the remaining 20\% (98
samples) were assigned to the test set. Using 100 splits of the data
accounts for the variation that may be observed depending on the samples
that are in the training or test sets. Each samples was in the training
set an average of N times and the test set and average of N times.

The data was processed through two pathways. First, the standard pathway
using the OptiClust algorithm. In this pathway, all of the data was
clustered together to generate OTUs and the resulting abundance tables
were split into the training and testing sets. In the second pathway,
the pre-processed data was split into the training and testing sets. The
training set was clustered into OTUs, then the test set was fit to the
OTUs of the training set using the OptiFit algorithm. The OptiFit
algorithm was run with method open so that any sequences that didn't map
to the existing OTU clusters would form new OTUs. Any OTUs that were not
in the training set were removed prior to machine learning. For both
pathways all data was sub-sampled to 10,000 reads per sample.

\textbf{\emph{Machine Learning.}} Machine learning using Random Forest
was conducted with the R package mikrompl (v XXXX) \{\} to predict the
diagnosis (SRN or normal) for the samples in the test set for each data
split. The training set was preprocessed to normalize values
(scale/center), collapse colinear features, and remove features with
zero-variance. The preprocessing from the training set was then applied
to the test set. P values comparing model performance were calculated as
previously described \{\}. The averaged ROC curves were plotted by
taking the average and standard deviation of the sensitivity at each
specificity value.

\textbf{\emph{Code Availability.}} all code used in this analysis is
available at
\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2021}

@@

(funding)

@@

\includegraphics{../exploratory/figures/figure1B.png}
\includegraphics{../exploratory/figures/figure1Apng.png}

@@

\includegraphics{../exploratory/figures/Figure2.png}
